---
title: "Error & Else"
author: "Wenxin Du"
date: "5/3/2018"
output: github_document
---
```{r global_options, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
```
```{r}
library(tidyr)
library(dplyr)
library(ggplot2)
library(lubridate)
library(readr)
library(readxl)
```
```{r}
JunSep16 <- read_excel("~/nomadic-herders/LTSdata/2016JunToSep.xls")
OctNov16 <- read_excel("~/nomadic-herders/LTSdata/2016OctNov.xlsx")
Apr17 <- read_excel("~/nomadic-herders/LTSdata/2017Apr.xlsx")
AugSep17 <- read_excel("~/nomadic-herders/LTSdata/2017AugSep.xlsx")
Dec17 <- read_excel("~/nomadic-herders/LTSdata/2017Dec.xlsx")
JanMar17 <- read_excel("~/nomadic-herders/LTSdata/2017JanToMar.xlsx")
Jul17 <- read_excel("~/nomadic-herders/LTSdata/2017Jul.xlsx")
Jun17 <-read_excel("~/nomadic-herders/LTSdata/2017Jun.xlsx")
May17 <- read_excel("~/nomadic-herders/LTSdata/2017May.xlsx")
OctDec17 <- read_excel("~/nomadic-herders/LTSdata/2017OctDec.xlsx")
X2016_Dec <- read_excel("~/nomadic-herders/LTSdata/2016 Dec.xlsx")
```


```{r}
JunSep16 <- JunSep16 %>%
  mutate(Number = as.numeric(substr(Number, 2, 100)))
OctNov16 <- OctNov16 %>%
  mutate(Number = as.numeric(Number), Name = as.numeric(Name))
Jul17 <- Jul17 %>%
  mutate(Name = as.numeric(Name))
LTS <- bind_rows(Apr17, AugSep17, Dec17, JanMar17, Jul17, Jun17, JunSep16, May17, OctDec17, OctNov16, X2016_Dec)
LTS <- select(LTS, Date, Time, Type, Number, Message)
```
```{r}
LTS <- LTS %>%
  mutate(Number = ifelse(nchar(as.character(Number)) == 11, Number %% 100000000, Number),
         Number = ifelse(nchar(as.character(Number)) == 10, Number %% 10000000, Number))
```
```{r}
n <- LTS %>%
  distinct(Number)
nl <- as.vector(n$Number)
nr <- n %>%
  nrow()
```
```{r}
set.seed(111)
id_ <- sample(10000:99999, nr, replace=FALSE)
id <- floor(runif(nr, min = 10000, max = 99999)) ###wrong code, not truly distinct numbers generated
```
```{r}
identifier <- data.frame(nl, id_) %>%
  mutate(Number = nl, id = id_) %>%
  select(Number, id)
```
```{r}
LTS_deidentified_1_ <- read_csv("~/nomadic-herders/data/LTS_deidentified (1).csv")
In_4_ <- read_csv("~/nomadic-herders/LTSdata/In (4).csv")
LTS_yearmonth <-  LTS_deidentified_1_ %>%
  mutate(Date = ymd(Date)) %>%
  mutate(Year = year(Date), month = month(Date)) %>%
  mutate(month = ifelse(nchar(as.character(month)) == 1, paste("0", sep = "", as.character(month)), as.character(month))) %>%
  mutate(year_month = paste(as.character(Year), "-", month)) 
```


### Distribution of Unique Users over Time

```{r}
LTS_yearmonth %>%
  group_by(year_month) %>%
  distinct(id) %>%
  summarize(n = n()) %>%
  arrange(desc(n)) %>%
  ggplot(aes(x = year_month, y = n)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 40, hjust = 1)) +
  ggtitle("Number of Unique Users over Time") +
  xlab("Year-Month") +
  ylab("Number of Unique users")
```

### Distribution of Odd Telephone Numbers Over Time
#### Odd telephone numbers: phone numbers less than 8 digits

```{r}
odd_phone_numbers <- left_join(LTS, identifier) %>%
  filter(nchar(as.character(Number)) != 8) %>%
  distinct(id) %>%
  mutate(is_odd = TRUE)
a <- full_join(LTS_yearmonth, odd_phone_numbers) %>%
  mutate(is_odd = ifelse(!is.na(is_odd), TRUE, FALSE))
a %>%
  group_by(year_month, is_odd) %>%
  distinct(id) %>%
  summarize(n = n()) %>%
  ggplot(aes(x = year_month, y = n, fill = is_odd)) + 
  geom_col()+
  theme(axis.text.x = element_text(angle = 40, hjust = 1))

a %>% 
  group_by(year_month, is_odd) %>%
  distinct(id) %>%
  summarize(n = n()) %>%
  mutate(prop = n/sum(n)) %>%
  ggplot(aes(x = year_month, y = prop, fill = is_odd)) + 
  geom_col()+
  theme(axis.text.x = element_text(angle = 40, hjust = 1)) +
  scale_fill_manual(values = c("white", "gray"))
```

### Create a list of IDs who sent in requests but never hear back from the system

```{r}
never_hear_back <- LTS_yearmonth %>%
  group_by(id, Type) %>%
  summarize(n = n()) %>%
  spread(key = Type, value = n) %>%
  filter(is.na(out)) %>%
  arrange(desc(`in`)) 
nhb <- as.vector(never_hear_back$id)
```

```{r}
In <- In_4_ %>%
  mutate(date = ymd(Date)) %>%
  select(-Date) %>%
  rename(Date = date)
```

### Distribution of messages sent in messages with correct area code and correct format but never hear back over time

```{r}
bb <- In %>%
  mutate(Date = ymd(Date)) %>%
  mutate(Year = year(Date), month = month(Date)) %>%
  mutate(month = ifelse(nchar(as.character(month)) == 1, paste("0", sep = "", as.character(month)), as.character(month))) %>%
  mutate(year_month = paste(as.character(Year), "-", month)) 
```

```{r}
pattern <- "^[0-9]{5}\\s+[0-9]{1}$"
bb <- bb %>%
  mutate(correct = grepl(pattern, Message))
bb %>%
  group_by(correct) %>%
  summarize(n = n())
```

```{r}
bb %>%
  group_by(year_month) %>%
  filter(id %in% nhb) %>%
  filter(correct == TRUE) %>%
  filter(area_correct == "real") %>%
  group_by(area_correct, year_month) %>%
  summarize(n = n()) %>%
  ggplot(aes(x = year_month, y = n)) + 
  geom_col(position = "dodge") +
  theme(axis.text.x = element_text(angle = 40, hjust = 1))
```

### Distribution of messages with area code not existing

```{r}
bb %>%
  group_by(year_month) %>%
  filter(area_correct == "fake") %>%
  group_by(area_correct, year_month) %>%
  summarize(n = n()) %>%
  ggplot(aes(x = year_month, y = n)) + 
  geom_col(position = "dodge") +
  theme(axis.text.x = element_text(angle = 40, hjust = 1))
```

### Distribution of messages with area code not existing from users who never hear back

```{r}
bb %>%
  group_by(year_month) %>%
  filter(id %in% nhb) %>%
  filter(area_correct == "fake") %>%
  group_by(area_correct, year_month) %>%
  summarize(n = n()) %>%
  ggplot(aes(x = year_month, y = n)) + 
  geom_col(position = "dodge") +
  theme(axis.text.x = element_text(angle = 40, hjust = 1))
```

### Distribution of messages from those who never hear back

```{r}
bb %>%
  group_by(year_month) %>%
  filter(id %in% nhb) %>%
  filter(correct == TRUE) %>%
  group_by(area_correct, year_month) %>%
  summarize(n = n()) %>%
  ggplot(aes(x = year_month, y = n, fill = area_correct)) + 
  geom_col(position = "dodge") +
  theme(axis.text.x = element_text(angle = 40, hjust = 1))
```

### Distribution of number of requests for top 6 most popular area codes

```{r}
area <- bb %>%
  filter(area_correct == "real", !is.na(area)) %>%
  group_by(area) %>%
  summarize(n = n()) %>%
  arrange(desc(n)) %>%
  head(6)
area
```
```{r}
qwq <- as.vector(area$area)
qqq <- bb %>%
  mutate(area = as.character(area)) %>%
  group_by(area, area_correct, year_month) %>%
  filter(area %in% qwq) %>%
  filter(area_correct == "real") %>%
  summarize(n = n()) %>%
  arrange(desc(n))
```
```{r}
qqq$code = factor(qqq$area, levels = c(62267, 84217, 46135, 67179, 23177, 62110))
qqq%>%
  ggplot(aes(x = year_month, y = n)) + geom_col(position = "dodge") +
  facet_wrap(~code, ncol = 2) +
  theme(axis.text.x = element_text(angle = 70, hjust = 1))+ 
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
    panel.background = element_blank(), axis.line = element_blank())
```

### Distribution of New Users

```{r}
new_users <- bb %>%
  group_by(id) %>%
  summarize(min = min(Date)) %>%
  mutate(Year = year(min), month = month(min)) %>%
  mutate(month = ifelse(nchar(as.character(month)) == 1, paste("0", sep = "", as.character(month)), as.character(month))) %>%
  mutate(year_month = paste(as.character(Year), "-", month))%>%
  group_by(year_month) %>%
  summarize(n = n()) 
new_users %>%
  ggplot(aes(x = year_month, y = n)) + geom_col() +
  theme(axis.text.x = element_text(angle = 40, hjust = 1))
```

### Distribution of proportion of new users each month

```{r}
bb %>%
  group_by(year_month) %>%
  distinct(id) %>%
  summarize(nid = n()) %>%
  full_join(new_users, by = "year_month") %>%
  mutate(prop = n/nid) %>%
  ggplot(aes(x = year_month, y = prop)) + geom_col() + theme(axis.text.x = element_text(angle = 40, hjust = 1))
```

### Distribution of new users each month (includes whether they dropped within the same month)

```{r}
gg <- bb %>%
  group_by(id) %>%
  summarize(min = min(Date), max = max(Date)) %>%
  mutate(Yearmin = year(min), monthmin = month(min), Yearmax = year(max), monthmax = month(max)) %>%
  mutate(monthmin = ifelse(nchar(as.character(monthmin)) == 1, paste("0", sep = "", as.character(monthmin)), as.character(monthmin))) %>%
  mutate(monthmax = ifelse(nchar(as.character(monthmax)) == 1, paste("0", sep = "", as.character(monthmax)), as.character(monthmax))) %>%
  mutate(year_month_min = paste(as.character(Yearmin), "-", monthmin))%>%
  mutate(year_month_max = paste(as.character(Yearmax), "-", monthmax))%>%
  mutate(drop_same_month = year_month_min == year_month_max) %>%
  group_by(year_month_min, drop_same_month) %>%
  summarize(n = n()) 
gg %>%
  ggplot(aes(x = year_month_min, y = n, fill = drop_same_month)) + geom_col() +
  theme(axis.text.x = element_text(angle = 40, hjust = 1))
```

```{r}
gg %>%
  filter(year_month_min != "2017 - 12") %>%
  group_by(year_month_min) %>%
  mutate(prop = n/sum(n)) %>%
  ggplot(aes(x = year_month_min, y = prop, fill = drop_same_month))+geom_col()+
  theme(axis.text.x = element_text(angle = 40, hjust = 1))
```

### Distribution of Those occured only once

```{r}
c <- LTS_yearmonth %>%
  group_by(id) %>%
  summarize(n = n()) %>%
  filter(n == 1) 
d <- as.vector(c$id)
LTS_yearmonth %>%
  mutate(only_once = ifelse(id %in% d, TRUE, FALSE))%>%
  mutate(out_only_once = ifelse(only_once == TRUE & Type == "out", TRUE, FALSE)) %>%
  group_by(year_month, only_once) %>%
  distinct(id) %>%
  summarize(n = n()) %>%
  arrange(desc(n)) %>%
  ggplot(aes(x = year_month, y = n, fill = only_once)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 40, hjust = 1))
```


```{r}
LTS_yearmonth %>%
  mutate(only_once = ifelse(id %in% d, TRUE, FALSE))%>%
  mutate(out_only_once = ifelse(only_once == TRUE & Type == "out", TRUE, FALSE)) %>%
  group_by(year_month, only_once) %>%
  distinct(id) %>%
  summarize(n = n()) %>%
  group_by(year_month) %>%
  mutate(prop = n/sum(n)) %>%
  ggplot(aes(x = year_month, y = prop, fill = only_once)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 40, hjust = 1)) 
```
```{r}
LTS_yearmonth %>%
  mutate(only_once = ifelse(id %in% d, TRUE, FALSE))%>%
  mutate(out_in = ifelse(only_once == TRUE & Type == "out", "ONCE OUT", ifelse(only_once == TRUE & Type == "in", "ONCE IN", "MORE THAN ONCE"))) %>%
  group_by(year_month, out_in) %>%
  distinct(id) %>%
  summarize(n = n()) %>%
  group_by(year_month) %>%
  mutate(prop = n/sum(n)) %>%
  ggplot(aes(x = year_month, y = prop, fill = out_in)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 40, hjust = 1)) +
  scale_fill_manual(values=c("mistyrose2", "lightskyblue", "lightcoral"))
```



